---
title: "PA1_template"
author: "Lin CK"
date: "Sunday, May 17, 2015"
output: html_document 
---

# Reproducible Research Assessment 1  

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Prepare the R environment  

Throughout this report when writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code.  

First, we set echo equal a TRUE and results equal a 'hold' as global options for this document.  
```{r}
library(knitr)
require(markdown) ## require(markdown) to create md file
opts_chunk$set(echo = TRUE, results = 'hold')
```

### Load the required libraries  

```{r}
library(data.table)
library(ggplot2) # we shall use ggplot2 for plotting figures
```

## Loading and preprocessing the data  

1. Load the data.  
2. Process the data into a format suitable for your analysis  

### Load the required data  
Use read.csv(). It is assumed that the file activity.csv is saved in the current working directory.  
```{r}
rdata <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
```

### Tidying the data  

Convert the date field to Date class and interval to Factor class.  
```{r}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
```

Check the data using str() method:
```{r}
str(rdata)
```  

## What is the mean total number of steps taken per day?

Ignore the missing values, we calculate the total steps per day.  

```{r}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```  

1. Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interval.

```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```  

2. Calculate the mean and median of the number of steps taken per day.  

```{r}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```  
The mean is `r steps_mean` and median is `r steps_median`. 

## What is the average daily activity pattern?  

We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.  

```{r}
steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
##this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```  

1. We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals.

```{r}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```

2. Now, we find the 5-minute interval with the containing the maximum number of steps:


```{r}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
```

The `r max_interval` has the maximum number of steps.  

## Inputing missing values:

### 1. Total number of missing values:

The total number of missing values:

```{r}
missing_vals <- sum(is.na(rdata$steps))
```

The total number of missing values are `r missing_vals`.  

### 2. Strategy for filling in all of the missing values in the dataset  

Create a na_fill(data, pervalue)  

```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
```  

Check any remaining missing values  

```{r}
any_missingvalues <- sum(is.na(rdata_fill$steps))
```  

Output shows that there are `r any_missingvalues` missing values remaining.  

### 3. A histogram of the total number of steps taken each day

Plot a histogram  

```{r}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

### Calculate the mean and median

```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

The mean is `r steps_mean_fill` and median is `r steps_median_fill`.

### Do these values differ from the estimates in the first part of the assignment?  

Yes, these values do differ slightly.  

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

As you can see, comparing with the calculations done in the first section of this document, we observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean.

Since our data has shown a t-student distribution (see both histograms), it seems that the impact of imputing missing values has increase our peak, but it's not affect negatively our predictions.

## Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.
1. Augment the table with a column that indicates the day of the week
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
3. Tabulate the average steps per interval for each data set.
4. Plot the two data sets side by side for comparison.

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```

See the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:  

```{r}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

We can see at the graph above that activity on the weekday has the greatest peak from all steps intervals. But, we can see too that weekends activities has more peaks over a hundred than weekday. This could be due to the fact that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport. In the other hand, at weekend we can see better distribution of effort along the time.